%!TEX root = thesis.tex
\chapter{Background and Theory}
\label{chap:background}

This chapter covers the technical background for both studies in this thesis. It starts
with wind turbine physics and control, then moves to SCADA systems and condition
monitoring, the digital twin concept, and the machine learning methods used in the two
studies: rule-based classifiers, supervised classifiers, and deep reinforcement learning.
Readers already familiar with a topic can move to the next section.

% ============================================================
\section{Wind Turbine Aerodynamics and Control}
\label{sec:bg:aero}
% ============================================================

\subsection{Power Production and the Betz Limit}
\label{subsec:bg:betz}

A wind turbine captures energy from moving air. The power available in a stream of wind
is given by $P = 0.5 \times \rho \times A \times v^3$, where $\rho$ is the air density
(approximately 1.225\,kg/m$^3$ at sea level), $A$ is the rotor swept area, and $v$ is
the wind speed \citep{manwell2009}. The cubic relationship with wind speed means that
doubling the wind speed increases the available power eightfold. This is why the
transition between partial and full load operation matters so much: a turbine producing
500\,kW at 7\,m/s would theoretically need to handle 4,000\,kW at 14\,m/s, far beyond
what its generator can take.

Not all of this energy can be captured. If the turbine slowed the wind to a complete
stop, no more air could flow through the rotor. If it did nothing, no energy would be
taken out. The theoretical maximum fraction of wind power that can be captured is 16/27,
approximately 59.3\%. This is known as the Betz limit, named after Albert Betz who
derived it in 1919 \citep{manwell2009}. In practice, modern large wind turbines reach a
power coefficient ($C_p$) of around 0.45 to 0.50 at their best operating point
\citep{manwell2009}.

The power coefficient $C_p$ depends on two parameters: the tip speed ratio (TSR), which
is the ratio of the blade tip speed to the wind speed, and the blade pitch angle. Each
turbine has an optimal TSR at which $C_p$ is highest, and the control system works to
stay at this point during partial load operation. In full load, the goal changes from
maximising $C_p$ to keeping power output at its rated value, which means the turbine
deliberately reduces its aerodynamic efficiency by adjusting the blade pitch.

\subsection{Turbine Control Regions}
\label{subsec:bg:regions}

The operating range of a wind turbine is typically split into three main regions based on
wind speed. Below the cut-in wind speed (approximately 3--4\,m/s for most modern
turbines), the wind is too weak to produce useful power and the turbine stays in standby.
Between cut-in and rated wind speed (typically 12--13\,m/s), the turbine operates in
partial load, trying to capture as much energy as possible. Above rated wind speed up to
the cut-out wind speed (typically 25\,m/s), the turbine operates in full load, actively
limiting its power output to protect the drivetrain and generator from overload
\citep{manwell2009}.

These three regions are the foundation of the zone framework used in this thesis, but
they are not enough on their own. The transitions between regions, starting up, shutting
down, and the moment the generator engages, produce sensor readings that do not fit
cleanly into any of the three main regions. The nine-zone framework developed in this
thesis \citep{indrekvam2026} captures these transition states explicitly, which is
important for avoiding misclassification of normal transient behaviour as faults.

\subsection{Blade Pitch and Generator Torque Control}
\label{subsec:bg:control}

Modern variable-speed wind turbines use two main control mechanisms: the blade pitch
system and the generator torque controller \citep{manwell2009}. In partial load (Zone~7 in this thesis),
blade pitch is held fixed at approximately zero degrees, the position that gives the best
aerodynamic efficiency, while generator torque is adjusted to keep the tip speed ratio at
its optimal value as wind speed changes. This is called torque control.

In full load (Zone~8), the generator has reached its rated torque and cannot absorb more
power. The control system switches to pitch control: the blades are rotated to increase
their angle of attack with the wind, which reduces the aerodynamic force on the rotor and
keeps rotational speed within safe limits. The boundary between these two control modes
is one of the clearest features in the SCADA data and is central to the zone definitions
in Chapter~3.

Safety lockdown (Zone~1) is a separate state from normal shutdown. When a fault is
detected or a maintenance command is issued, the blades are immediately pitched to
approximately 90 degrees, the feathered position where they present the smallest possible
surface area to the wind, and the rotor is braked to a stop. This leaves a very clear
signature in the pitch angle sensor and makes Zone~1 one of the easiest zones to
identify.

% ============================================================
\section{SCADA Systems and Condition Monitoring}
\label{sec:bg:scada}
% ============================================================

\subsection{What SCADA Systems Measure}
\label{subsec:bg:scada:measure}

A wind turbine SCADA system records operational and environmental data at regular
intervals throughout the turbine's life. Most modern turbines store data as ten-minute
averages, alongside the minimum, maximum, and standard deviation of each variable over
the same period. A typical turbine records 50 to 100 variables, covering wind speed and
power output through to oil temperatures, bearing vibration, and grid frequency
\citep{pandit2024}. The ten-minute averaging interval is long enough to smooth out
short-term turbulent fluctuations while still being short enough to capture transitions
between operating states.

The four SCADA variables most central to this thesis are ambient wind speed
(\texttt{Amb\_Wind\_Speed}, measured by an anemometer on the nacelle), rotor angular
speed (\texttt{Rtr\_RPM\_Avg}, measured by an encoder on the main shaft), blade pitch
angle (\texttt{Blades\_Pitch\_Angle}, the average pitch of all three blades), and grid
power output (\texttt{Grid\_Prod\_Power}, the net electrical power delivered to the
grid). These four variables together describe the main aerodynamic and electrical state
of the turbine. As shown in Chapter~4, a classifier trained on these four variables alone
is sufficient to identify all operational zones with 100\% accuracy.

\subsection{Closed-Loop Data and Its Challenges}
\label{subsec:bg:scada:closedloop}

A key challenge with SCADA data that is often overlooked in the machine learning
literature is that it comes from a closed-loop control system \citep{pandit2024}. The turbine does not
passively observe the wind and report what happens. It actively responds to changing
conditions through its pitch and torque controllers. This means that many SCADA variables
are outputs of the control system rather than independent measurements of a physical
process.

For example, at rated wind speed the power output is held nearly constant by the control
system, even though the available wind power varies significantly around that point. An
analysis that treats power output as a simple function of wind speed will not capture
this behaviour. Similarly, blade pitch angle is a control output rather than an
environmental measurement: it tells us what the control system decided to do, not what
the wind was doing. Understanding the link between these control actions and the
turbine's operating zone is central to building useful SCADA-based models.

The zone classification approach in this thesis directly addresses this challenge. By
identifying which control regime the turbine is in, the zone label gives the context
needed to interpret every other SCADA variable correctly. A power reading of 800\,kW in
Zone~7 (partial load) follows the expected cubic power curve and indicates normal
operation. The same reading in Zone~8 (full load) would indicate significant
underperformance, because the turbine should be producing its rated power in that zone.

\subsection{Condition Monitoring Using SCADA}
\label{subsec:bg:scada:cm}

Condition monitoring means continuously tracking a system's health using sensor data,
with the goal of catching faults or degradation before they cause a failure. For wind
turbines, SCADA-based condition monitoring is attractive because the data already exists
and is collected automatically, with no extra instrumentation required \citep{pandit2024}.
Physics-informed approaches to SCADA data preprocessing, such as using physical
constraints to identify and remove anomalous readings, have been shown to improve the
reliability of models built on this data \citep{hong2025}.

The basic idea is to build a model of what the turbine's sensor readings look like under
normal conditions, then flag deviations from this model as potential faults. This
requires the model to know which operating zone the turbine is currently in: a power
reading that looks wrong in one zone may be perfectly normal in another. Without zone
classification, a condition monitoring system cannot tell the difference between a fault
and a normal transition between operating states.

The time between fault initiation and eventual failure varies a lot across different
failure modes. Blade surface erosion develops over months or years. Generator bearing
faults may progress over weeks. Electrical faults can cause failure in seconds. A good
condition monitoring system needs to work across all of these timescales, which requires
both short-term monitoring and longer-term trend analysis, both of which depend on
correct zone classification to avoid false alarms.

\subsection{Power Curve Modelling}
\label{subsec:bg:scada:powercurve}

The wind turbine power curve describes the relationship between wind speed and power
output and is one of the most widely used tools in wind turbine performance monitoring
\citep{pandit2019}. A turbine producing less power than expected for a given wind speed
is either experiencing a fault, degradation, or unfavourable conditions such as high
turbulence. Power curve analysis is a standard method for detecting long-term performance
degradation.

However, a single power curve model covering all operating conditions will be inaccurate,
because the relationship between wind speed and power output changes fundamentally between
zones. In partial load (Zone~7), power follows a smooth cubic curve with wind speed. In
full load (Zone~8), power is nearly constant regardless of wind speed. In Zone~6,
generator engagement introduces a sharp nonlinearity. A realistic power curve model for
PHM therefore needs a separate model for each zone. This is only possible once the zones
have been correctly identified in the data.

% ============================================================
\section{Digital Twins in Prognostics and Health Management}
\label{sec:bg:dt}
% ============================================================

\subsection{Definitions and Concepts}
\label{subsec:bg:dt:def}

A digital twin is a computational model of a physical asset that is continuously updated
with real sensor data. The definition adopted in this thesis comes from the National
Academies of Sciences, Engineering, and Medicine \citeyearpar{nationalacademies2024}:
\textit{``A set of virtual information constructs that mimics the structure, context, and
behaviour of a natural, engineered, or social system; is dynamically updated with data
from its physical counterpart; possesses predictive capabilities; and informs decisions
to realize value.''}

For a wind turbine, a digital twin might include models of the power curve, the
drivetrain dynamics, structural loads on the tower and blades, and the expected thermal
behaviour of key components. Each model is updated regularly using new SCADA data so
that the digital version stays aligned with the real turbine as it ages. When a
significant gap appears between what the digital twin predicts and what the sensors
record, this signals a potential fault or degradation event.

A digital twin is not a single model but a collection of interacting models, each with a
different role and a different update frequency. Some are physics-based and change slowly
as physical parameters are recalibrated. Others are data-driven and can be retrained as
new data arrives. The zone classification models in this thesis act as a context layer
that determines which other models should be active at any given moment.

\subsection{Requirements for an Interpretable and Adaptive Digital Twin}
\label{subsec:bg:dt:req}

For a digital twin to be useful in a real PHM setting it needs to meet several practical
requirements. First, it should be interpretable: operators and maintenance engineers need
to understand why the system raised an alert, not just that an alert was raised. A
black-box model that produces accurate predictions but no explanation is hard to trust
in a safety-critical context. This is one of the reasons why rule-based models, despite
their lower accuracy compared to neural networks, remain attractive for industrial
applications.

Second, the digital twin should be adaptive. Turbine behaviour changes as the machine
ages: bearings wear gradually, blade surfaces erode, and pitch control mechanisms
develop tolerances. A static model trained once on historical data will slowly drift away
from the real turbine. The digital twin needs a way to update its models using new
operational data, ideally without requiring a human expert to re-label data every time a
model needs refreshing. Recent work on hybrid physics and data-driven models for wind
turbine power prediction shows that combining a physics submodel with a data-driven
residual model achieves significantly higher accuracy than either approach alone, while
remaining interpretable \citep{gijon2024}.

Third, the models in the digital twin need to be context-aware. A single global model of
turbine behaviour is less useful than a set of zone-specific models that each describe
the expected behaviour in one operating state. Operational zone classification is
therefore not just one task among many. It is the prerequisite for making every other
model in the digital twin context-aware. This thesis focuses specifically on this
prerequisite task.

% ============================================================
\section{Rule-Based Machine Learning}
\label{sec:bg:rules}
% ============================================================

\subsection{The RIPPER Algorithm}
\label{subsec:bg:ripper}

RIPPER (Repeated Incremental Pruning to Produce Error Reduction) is a rule induction
algorithm developed by \citet{cohen1995}. It learns a set of IF--THEN classification
rules from labelled training data. Each rule takes the form: IF (condition$_1$ AND
condition$_2$ AND \ldots) THEN class. The conditions refer to specific values or ranges
of discretized input features. RIPPER is a sequential covering algorithm: it learns rules
for one class at a time, removes the data points covered by those rules, then learns
rules for the next class from the remaining data.

RIPPER uses a minimum description length (MDL) criterion to balance accuracy against rule
complexity. After building each rule to fit the training data, a pruning step removes
conditions that do not improve performance on a held-out pruning set. This prevents
overfitting and tends to produce compact, readable rules. The main hyperparameters are
the pruning set size, the number of optimisation passes ($k$), and a description length
allowance that controls how much rule complexity is allowed.

In this thesis, RIPPER is used in a one-vs-rest setup: a separate binary RIPPER model is
trained for each of the eight zones present in the data, where the positive class is the
zone of interest and the negative class is all other zones. This produces eight
independent models, each outputting a probability that the current observation belongs to
its assigned zone. These probabilities are then used as input features for the DRL agents
and supervised classifiers in Chapter~4.

\subsection{The FOIL Algorithm}
\label{subsec:bg:foil}

FOIL (First-Order Inductive Learner) is a rule learning algorithm developed by
\citet{quinlan1990}. Like RIPPER, it produces IF--THEN classification rules. Unlike
RIPPER, FOIL was originally designed to learn first-order logic rules, though in this
thesis it is used in its standard propositional form where the rule structure is the same
as RIPPER.

FOIL builds rules by starting with an empty rule and adding conditions one at a time,
always choosing the condition that gives the biggest improvement in separating the target
class from the rest. It keeps adding conditions until the rule covers only positive
examples or no further improvement is possible. This approach tends to produce more
specific rules than RIPPER, with each rule covering a smaller part of the feature space,
which means FOIL usually generates more rules per zone.

When multiple FOIL rules match a given observation, a conflict resolution step is needed.
In this thesis, the prediction goes to the zone whose matching rule has the most
conditions, on the reasoning that a more specific rule is a better match for the
observation than a more general one. If no FOIL rule matches at all, the prediction falls
back to the RIPPER model output. This strategy is fixed before training and applied the
same way during both internal and external validation.

\subsection{Strengths and Limitations of Rule-Based Methods}
\label{subsec:bg:rules:limits}

The main strength of rule-based methods is that the learned rules can be read and checked
by a domain expert. A rule that says ``IF rotor RPM is in the torque-controlled range AND
wind speed is medium THEN Zone~7'' can be verified against turbine physics. This makes
rule-based models suitable for safety-critical applications where a human operator needs
to understand and trust the classification.

A key limitation is that rule-based methods require discretized inputs. Continuous sensor
values like wind speed or rotor RPM must be converted into discrete bins before the rules
can be learned. This introduces binning error: observations near a bin boundary may
end up in the wrong bin and be classified incorrectly. As shown in Chapter~4, this is the
main reason why RIPPER and FOIL reach around 91--92\% accuracy while a Random Forest
using the same four continuous features achieves 100\%. Despite this limitation,
rule-based methods remain an active area of research. \citet{nossig2024} recently showed
that RIPPER and FOIL can scale to large, high-dimensional datasets using a modularity
approach, keeping accuracy close to black-box models while preserving full
interpretability.

A second limitation is that rule-based models are static. Once trained, they do not
update when the turbine's behaviour changes. New rules must be learned from scratch on
new labelled data. For a PHM digital twin expected to track changes in turbine behaviour
over years of operation, this is a practical drawback.

% ============================================================
\section{Supervised Machine Learning Classifiers}
\label{sec:bg:supervised}
% ============================================================

\subsection{Random Forests}
\label{subsec:bg:rf}

A Random Forest \citep{breiman2001} is an ensemble classifier that builds many decision
trees, each trained on a random sample of the training data and using a random subset of
features at each split. The final prediction is made by majority vote across all trees.
The combination of randomness in data sampling and feature selection makes each tree
slightly different, and the ensemble tends to be much more accurate and stable than any
individual tree.

Random Forests handle multiclass classification directly, are not sensitive to irrelevant
features, and need very little hyperparameter tuning. The most important setting is the
number of trees: more trees reduce variance but take longer to train, and performance
usually levels off beyond 100--200 trees. Random Forests also give a built-in measure of
how important each feature is, based on how often it is used in splits and how much those
splits reduce prediction error.

In this thesis, Random Forests are trained with 300 trees and balanced class weights to
account for the unequal distribution of samples across the eight zones. The balanced
setting adjusts the weight of each sample in proportion to the inverse frequency of its
class, which stops the model from being dominated by Zone~7, which accounts for
approximately 51\% of the training data.

\subsection{Histogram Gradient Boosting}
\label{subsec:bg:gbt}

Gradient Boosting \citep{friedman2001} builds an ensemble of decision trees one at a
time, where each new tree is trained to correct the errors made by all previous trees.
Unlike Random Forests where all trees are trained independently, Gradient Boosting trees
are trained in order and each one fits the residual errors of the current ensemble. This
makes Gradient Boosting very accurate but also more sensitive to hyperparameter choices.

The Histogram Gradient Boosting variant used in this thesis speeds up training by first
grouping continuous feature values into discrete bins before finding the best split at
each node. This reduces the number of candidate splits to check from $O(n)$ to
$O(\text{bins})$, which makes training much faster on large datasets without meaningfully
affecting accuracy. The implementation used is scikit-learn's
\texttt{HistGradientBoostingClassifier} with 300 iterations, a learning rate of 0.05,
and a maximum tree depth of 8.

\subsection{Multi-Layer Perceptrons}
\label{subsec:bg:mlp}

A Multi-Layer Perceptron (MLP) is a feedforward neural network with an input layer, one
or more hidden layers of neurons, and an output layer. Each neuron computes a weighted
sum of its inputs, applies a non-linear activation function (ReLU in this thesis), and
passes the result to the next layer. The network is trained using backpropagation with
the Adam optimiser \citep{kingma2014}.

A sufficiently large MLP can in principle approximate any continuous function. In
practice, the number of layers and neurons must be chosen to balance model capacity
against the risk of overfitting. In this thesis, both the MLP trained on the full
13-feature set and the one trained on raw SCADA only use two hidden layers of 256 neurons
each. Early stopping based on a held-out validation split is used to prevent overfitting.

Unlike Random Forests and Gradient Boosting, MLP classifiers in scikit-learn require
integer class labels when early stopping is active. In this thesis, the zone names
(\texttt{Zone\_1} through \texttt{Zone\_8}) are converted to integers before training and
converted back to zone names for evaluation. This is a technical implementation detail
with no effect on the results.

% ============================================================
\section{Deep Reinforcement Learning}
\label{sec:bg:drl}
% ============================================================

\subsection{The Reinforcement Learning Framework}
\label{subsec:bg:drl:framework}

Reinforcement learning (RL) is a way of learning through interaction \citep{raffin2021}. An agent observes
the current state of an environment, selects an action, receives a reward signal
indicating how good the action was, and then observes the new state. The goal is to learn
a policy, a mapping from states to actions, that maximises the total reward received over
time. Unlike supervised learning, the agent is not told the correct action for each
state. It has to discover which actions are good through trial and error.

The formal framework is the Markov Decision Process (MDP), defined by a tuple
$(S, A, P, R, \gamma)$, where $S$ is the set of possible states, $A$ is the set of
possible actions, $P$ is the transition probability function, $R$ is the reward function,
and $\gamma$ is a discount factor between 0 and 1 that controls how much the agent values
future rewards compared to immediate ones. The Markov property requires that the next
state depends only on the current state and action, not on any earlier history.

In deep reinforcement learning (DRL), the policy or value function is represented by a
neural network, which lets the agent handle large or continuous state spaces. The three
DRL algorithms used in this thesis, DQN, PPO, and A2C, each take a different approach to
learning this policy, described in the sections below.

\subsection{Deep Q-Networks (DQN)}
\label{subsec:bg:dqn}

DQN, introduced by \citet{mnih2015}, learns a Q-function that estimates the total reward
the agent can expect by taking each action in each state. The Q-function is represented
by a deep neural network that takes the current state as input and outputs a Q-value for
each possible action. The agent always picks the action with the highest Q-value.
Training updates the network weights to reduce the difference between the predicted
Q-value and the target Q-value, which is the immediate reward plus the discounted best
Q-value in the next state.

Two ideas make DQN stable in practice. The first is an experience replay buffer: instead
of updating the network right after each step, the agent stores transitions (state,
action, reward, next state) in a memory buffer and trains on random mini-batches drawn
from this buffer. This breaks the correlation between consecutive training examples. The
second is a target network: a separate copy of the Q-network whose weights are only
updated occasionally. Using this target network to compute training targets stops the
targets from shifting too fast, which significantly improves stability.

\subsection{Proximal Policy Optimisation (PPO)}
\label{subsec:bg:ppo}

PPO, introduced by \citet{schulman2017}, directly optimises the policy rather than
learning a Q-function. Policy gradient methods update the policy to increase the
probability of actions that led to high rewards. The problem with simple policy gradient
methods is that large updates can destabilise training, sometimes causing the policy to
collapse to a poor solution it cannot recover from.

PPO addresses this by putting a limit on how much the policy is allowed to change at
each training step. The limit works by comparing old and new action probabilities
using a parameter $\varepsilon$ (typically 0.1 to 0.2): if the change would push the
ratio of new to old probabilities outside $[1-\varepsilon,\, 1+\varepsilon]$, the
update is stopped. This makes PPO much more stable than earlier policy gradient
methods while remaining simple to implement.

\subsection{Advantage Actor-Critic (A2C)}
\label{subsec:bg:a2c}

A2C is a synchronous variant of the Asynchronous Advantage Actor-Critic (A3C) algorithm
\citep{mnih2016}. It maintains two neural networks: an actor that outputs action
probabilities for the current state, and a critic that estimates the value of the current
state, meaning the total reward expected from this point onwards. The actor is updated to
increase the probability of actions that did better than expected and decrease the
probability of actions that did worse. The difference between actual reward and the
critic's estimate is called the advantage.

The actor and critic typically share most of their network weights, with separate output
heads for the policy and the value function. A2C collects a batch of experience from a
single environment, then updates both networks before collecting the next batch. This is
simpler to implement than the asynchronous A3C and reaches similar performance in most
settings. In this thesis, A2C shows the weakest performance of the three DRL algorithms,
with a specific failure mode on Zone~8 (full load operation) that is discussed in
Chapter~4.

\subsection{DRL in Industrial and Energy Applications}
\label{subsec:bg:drl:industrial}

DRL has been applied to a growing range of industrial control and monitoring tasks. In
wind energy, the most common use is active power and pitch control: training DRL agents
to maximise energy output or reduce mechanical loads by directly controlling blade pitch
and generator torque. \citet{xie2023} showed that a DRL agent could match or beat
model-based controllers on this task while needing less detailed system knowledge.
\citet{didier2022} applied a similar approach to floating offshore wind turbines, using a
trust region policy optimisation agent to control blade pitch in the full-load operating
region, a more challenging setting where platform motion adds complexity beyond
fixed-foundation turbines.

The use of DRL for classification tasks, as opposed to control tasks, is less common. In
a classification setting, the environment is the sequence of sensor observations, the
actions are class predictions, and the reward is a signal indicating whether the
prediction was correct. This is not the typical use of DRL, but it has one useful
property: the reward signal can in principle be defined without labelled data, using
physics rules or other automatic criteria instead. This is the motivation for applying
DRL to zone classification in this thesis, even though supervised classifiers achieve
higher accuracy in this study.

Stable-Baselines3 \citep{raffin2021}, the DRL library used in this thesis, provides
well-tested implementations of DQN, PPO, and A2C. The library is built on PyTorch and
uses the same commands for all algorithms that makes it straightforward to compare different
algorithms on the same environment.
