%!TEX root = thesis.tex
\chapter{Rule-Based Operational Zone Classification}
\label{chap:ch3}

\begin{quote}
\small
This chapter is based on the following publication:\\[4pt]
\textit{Indrekvam, G., Singh, M., Kampen, A.-L., Jha, M.~S. (2026).
Interpretable Operational Regime Classification for a Wind Turbine PHM Digital Twin
Architecture. Proceedings of the 9th European Conference of the Prognostics and Health
Management Society (PHME~2026).}\\[4pt]
The chapter is presented here in its thesis form, which is slightly expanded from the
conference paper to include additional methodological detail. The numerical results
presented here reflect the reimplemented models used consistently across both studies;
minor numerical differences from the conference paper are a consequence of this
methodological standardisation.
\end{quote}

\section{Motivation and Aim}
\label{sec:ch3:motivation}

The starting point for this study is a practical problem: wind turbine SCADA data cannot
be analysed reliably without knowing which operating state the turbine is in. A power
curve model that mixes data from partial load and full load will be inaccurate for both.
A fault detection system that cannot tell the difference between a zone transition and a
genuine anomaly will generate constant false alarms. The aim of this study is to build a
framework that solves this problem in a way that is both accurate and interpretable.

Interpretability is a specific requirement here, not just a nice-to-have. The zone
definitions must be explainable to a turbine operator or maintenance engineer who may not
have a machine learning background. If the classification rules cannot be read and checked
by a domain expert, it is very difficult to verify that they are correct or to diagnose
errors when the classifier makes a mistake. Rule-based methods are chosen specifically
because they produce readable IF--THEN rules that can be checked against known turbine
physics.

A second aim of this study is to produce a set of physics-informed zone labels that can
serve as training data for the machine learning models in Chapter~4. The zone framework
is therefore both the output of this study and an input to the next. The quality of the
zone labels directly determines the quality of every model trained on them.

\section{Data}
\label{sec:ch3:data}

This study uses SCADA data provided by EDP from Wind Turbine~7 of an offshore wind farm
located off the west coast of Africa. The turbine is a horizontal-axis offshore wind
turbine with a rated power of approximately 2\,MW. Data is available at ten-minute
resolution for two calendar years: 2016 and 2017. A total of 76 SCADA variables are
recorded, covering mechanical, electrical, and environmental measurements.

For the purposes of this study, four variables are used: ambient wind speed
(\texttt{Amb\_Wind\_Speed}, m/s), rotor angular speed (\texttt{Rtr\_RPM\_Avg}, RPM),
blade pitch angle (\texttt{Blades\_Pitch\_Angle}, degrees), and grid power output
(\texttt{Grid\_Prod\_Power}, kW). These four variables together capture the complete
aerodynamic and electrical state of the turbine. All other SCADA variables are excluded
from the classification models, though they remain available for downstream condition
monitoring tasks that build on the zone labels.

The 2016 dataset is used for training and internal validation, split chronologically:
the first 80\% of the time series (approximately 41,860 samples) forms the training set
and the remaining 20\% (approximately 10,466 samples) forms the internal validation set.
The full 2017 dataset (52,234 samples), collected from Wind Turbine~11 at the same wind
farm, serves as the external test set to measure how well the trained models generalise
to a different year and turbine. No 2017 data is used during training or validation. The
chronological split is used rather than a random split to respect the time-series
structure of the data and to test generalisation forward in time, which is the direction
that matters for operational deployment.

\section{Physics-Informed Zone Definitions}
\label{sec:ch3:zones}

The nine operational zones are defined from first principles of turbine control and
aerodynamics. Each zone corresponds to a physically well-defined state identified using
threshold rules applied to the four main SCADA variables. The threshold values are
derived from the known control parameters of the turbine rather than from statistical
analysis of the data. This means the zone boundaries are fixed before looking at the
SCADA measurements, which avoids any risk of the labelling procedure overfitting to the
specific dataset.

\subsection{Lockdown and Transition Regimes (Zones 1 and 2)}
\label{subsec:ch3:zones:12}

\textbf{Zone~1 --- Safety Lockdown}
(\texttt{Blades\_Pitch\_Angle} $\geq$ 86\textdegree): the turbine is intentionally
prevented from producing power. The blade pitch angle is held at approximately
90\textdegree, the feathered position where the blades present the smallest possible
aerodynamic surface to the wind. This state is typically triggered by external constraints
such as grid saturation or scheduled maintenance. The pitch threshold of 86\textdegree\
is conservative: any reading above this value unambiguously indicates lockdown regardless
of wind speed or rotor speed.

\textbf{Zone~2 --- Transition to or from Safety Lockdown}
(\texttt{Blades\_Pitch\_Angle} $<$ 86\textdegree, in transition): blade pitch is
actively adjusted as the turbine moves between normal operating zones and the lockdown
state. Transitions into or out of lockdown span more than one ten-minute SCADA interval,
producing intermediate pitch readings in this zone. Zone~2 is one of the most important
zones to identify correctly because misclassifying a transition as normal operation
produces misleading sensor readings in any model that assumes steady-state conditions.

\subsection{Standby and Transition Below Cut-In Wind Speed (Zones 3 and 4)}
\label{subsec:ch3:zones:34}

\textbf{Zone~3 --- Standby (Low-Wind Shutdown)}
(\texttt{Amb\_Wind\_Speed} $<$ 4\,m/s, \texttt{Blades\_Pitch\_Angle} $=$ 24\textdegree,
\texttt{Rtr\_RPM} $=$ 0): when wind speed is below cut-in, the turbine is fully braked.
The rotor does not move and blade pitch is fixed at approximately 24\textdegree. This is
a clearly defined state that is easy to identify from the rotor speed reading alone.

\textbf{Zone~4 --- Transition to or from Standby}
(\texttt{Amb\_Wind\_Speed} $<$ 4\,m/s, \texttt{Blades\_Pitch\_Angle} $\approx$
24\textdegree, \texttt{Rtr\_RPM} $>$ 0): during startup or shutdown at low wind speeds,
the brakes are released and the rotor begins to move, but wind conditions are not yet
enough for power production. This transient state is identified by a non-zero rotor speed
combined with low wind speed and a standby-range pitch angle. It is short-lived and
appears briefly when the turbine starts or stops under marginal wind conditions.

\subsection{Transitions to Power Production (Zones 5 and 6)}
\label{subsec:ch3:zones:56}

\textbf{Zone~5 --- Free Wheeling (Below Cut-In)}
(\texttt{Amb\_Wind\_Speed} $<$ 4\,m/s,
0\textdegree $<$ \texttt{Blades\_Pitch\_Angle} $<$ 24\textdegree,
0 $<$ \texttt{Rtr\_RPM} $<$ 11): as the turbine prepares to engage the generator,
blade pitch moves toward zero degrees to maximise aerodynamic efficiency. The rotor spins
slowly but the generator is not yet connected to the grid. This free-wheeling state can
also occur when wind speed drops below cut-in while the turbine is shutting down.

\textbf{Zone~6 --- Generator Engagement (Below Cut-In)}
(\texttt{Amb\_Wind\_Speed} $<$ 4\,m/s,
\texttt{Blades\_Pitch\_Angle} $=$ 0\textdegree,
11.0 $<$ \texttt{Rtr\_RPM} $\leq$ 11.2): at approximately 11.0--11.2\,RPM, the rotor
has reached the speed at which the generator can be connected to the grid. This zone
occupies a very narrow RPM band and is one of the most challenging to classify correctly.
Small measurement errors or data averaging effects can shift observations across the
Zone~5/6 or Zone~6/7 boundary. Zone~6 also sees the highest error rates from both
rule-based classifiers.

\subsection{Partial Load and Full Load Operation (Zones 7 and 8)}
\label{subsec:ch3:zones:78}

\textbf{Zone~7 --- Partial Load Operation}
(4 $\leq$ \texttt{Amb\_Wind\_Speed} $\leq$ 12\,m/s,
\texttt{Blades\_Pitch\_Angle} $\approx$ 0\textdegree,
11.2 $<$ \texttt{Rtr\_RPM} $<$ 14.8): the most common operating state, accounting for
approximately 51\% of the training data. The turbine produces power below its rated
capacity. Blade pitch is held at zero degrees to maximise aerodynamic efficiency and
power increases approximately as the cube of wind speed. The control system adjusts
generator torque to maintain the optimal tip speed ratio. Both RIPPER and FOIL achieve
high accuracy in Zone~7 due to the stability of this regime.

\textbf{Zone~8 --- Full Load Operation}
(12 $<$ \texttt{Amb\_Wind\_Speed} $<$ 25\,m/s,
\texttt{Blades\_Pitch\_Angle} $>$ 0\textdegree,
11.8 $\leq$ \texttt{Rtr\_RPM} $<$ 15): above rated wind speed, the turbine maintains
its maximum power output by progressively pitching the blades to reduce aerodynamic
efficiency. Rotor speed is held nearly constant at approximately 15\,RPM and power at
approximately 2000\,kW. The non-zero blade pitch combined with high wind speed is a
clear signature that makes Zone~8 straightforward to identify. Zone~8 accounts for
approximately 5.5\% of the validation data.

\textbf{Zone~9 --- High-Wind Safety Shutdown}
(\texttt{Amb\_Wind\_Speed} $>$ 25\,m/s): at extreme wind speeds the turbine shuts down
automatically to prevent structural overload. Blades are pitched toward 90\textdegree\
and the rotor is braked to a stop. No observations corresponding to Zone~9 were found in
the dataset. The classification problem therefore covers Zones~1--8 in practice.

\section{Discretization of SCADA Variables}
\label{sec:ch3:discretization}

RIPPER and FOIL operate on discrete (categorical) input features rather than continuous
values. The four SCADA variables are therefore converted to categorical bins before
training. The bin boundaries are set using the same physical thresholds that define the
zone boundaries, keeping the discretization consistent with the zone definitions.

Wind speed (\texttt{Amb\_Wind\_Speed}) is divided into four bins: below cut-in
($<$\,4\,m/s), medium wind (4--12\,m/s), high wind (12--25\,m/s), and above cut-out
($>$\,25\,m/s). Rotor speed (\texttt{Rtr\_RPM\_Avg}) is divided into five bins:
stationary (0\,RPM), low speed (0--11\,RPM), generator engagement (11--11.2\,RPM),
torque-controlled (11.2--14.8\,RPM), and rated ($\geq$\,14.8\,RPM). Blade pitch
(\texttt{Blades\_Pitch\_Angle}) is divided into five bins: full efficiency angle
($\approx$\,0\textdegree), standby angle ($\approx$\,24\textdegree), control angle
(between standby and lockdown transition), lockdown transition, and lockdown
($\geq$\,86\textdegree). Power output (\texttt{Grid\_Prod\_Power}) is divided into five
bins from very low to above rated.

The discretization boundaries were fixed before training and remained unchanged for the
2016 training/validation sets and the 2017 external test. This is important for fair
evaluation: the discretization is part of the model definition, not a parameter tuned to
the test data. The same boundaries are also used to build the RIPPER one-vs-rest
probability features in Chapter~4.

\section{Rule-Based Learning Framework}
\label{sec:ch3:framework}

\subsection{RIPPER Training}
\label{subsec:ch3:ripper}

RIPPER is set up as eight binary one-vs-rest classifiers, one for each zone present in
the data. Each classifier is trained to predict whether a given observation belongs to
one specific zone (the positive class) or any other zone (the negative class). This
produces eight independent RIPPER models, each with its own IF--THEN rules for its
assigned zone. The final prediction is the zone whose RIPPER model produces the highest
positive-class probability.

The RIPPER hyperparameters are: pruning set size $= 0.1$ (10\% of training data held
back for pruning), $k = 2$ (two optimisation passes after initial rule induction), and
description length allowance $= 64$. These values were chosen to produce rules that are
both accurate and compact. A larger pruning set gives simpler rules with slightly lower
training accuracy but better generalisation. A higher $k$ gives more optimised rules at
the cost of longer training time.

\subsection{FOIL Training and Rule Arbitration}
\label{subsec:ch3:foil}

FOIL is trained on the same discretized features as RIPPER. Unlike RIPPER, FOIL is
applied as a standard multiclass classifier rather than a one-vs-rest ensemble. For each
zone, FOIL generates rules by iteratively adding conditions that maximise information
gain until the rule covers only positive examples or no further improvement is possible.

FOIL may generate multiple rules that match a single observation. When this happens, the
prediction goes to the zone whose matching rule has the most conditions, on the reasoning
that a more specific rule is a better match than a more general one. If no FOIL rule
matches at all, the prediction falls back to the RIPPER model output. This arbitration
strategy is fixed before training and applied the same way during both internal and
external validation.

\subsection{Validation and Evaluation Protocol}
\label{subsec:ch3:eval}

The 2016 training set is used to train both RIPPER and FOIL. After training, both models
are evaluated on the 2016 validation set (the held-out last 20\% of 2016 data) without
any retraining or parameter adjustment. This internal validation measures how well the
models work on data from the same year and turbine as the training data.

Both models are then applied, again without retraining, to the complete 2017 dataset.
This external validation tests whether the learned rules remain valid one year later,
under potentially different weather patterns, with a turbine that has aged by one year.
The comparison between internal and external performance is the primary indicator of
whether the zone framework captures stable physical behaviour or year-specific patterns.

Performance is measured using overall accuracy and macro-averaged F1 score. The macro-F1
gives equal weight to all eight zones in the data, making it sensitive to errors in small
zones such as Zone~2 (97 samples in the validation set). Per-zone precision and recall
are also reported to identify which zones are most difficult to classify correctly.

\section{Results}
\label{sec:ch3:results}

\subsection{Internal Validation (2016)}
\label{subsec:ch3:results:val}

On the 2016 internal validation set, RIPPER achieves an overall accuracy of 92.6\% with
a macro-F1 of 0.916. The results show strong performance across most zones. The largest
errors occur in Zone~5 (free wheeling), where RIPPER achieves a precision of 0.62 and
recall of 0.98. This means RIPPER correctly identifies most Zone~5 observations but also
pulls in many observations from neighbouring zones, Zone~6 in particular. Zone~6 itself
shows reduced recall (0.71), reflecting the difficulty of separating the narrow generator
engagement band from adjacent zones using discretized bin features.

FOIL achieves 92.3\% accuracy and a macro-F1 of 0.897 on the same validation set. Its
errors are distributed differently from RIPPER's. Zone~4 (transition to or from standby)
shows reduced precision (0.75) and recall (0.94), and Zone~5 shows a recall of only 0.46
despite a precision of 1.00. This means FOIL is very conservative about predicting
Zone~5: it correctly identifies the few Zone~5 observations it does predict, but misses
more than half of them. This reflects the more specific nature of FOIL rules, which cover
a smaller region of the feature space and therefore miss more observations near zone
boundaries.

Both classifiers perform very well on the most common and clearly defined zones. Zone~1
(safety lockdown) achieves perfect or near-perfect precision and recall for both
classifiers, reflecting the unambiguous blade pitch signature. Zone~3 (standby) and
Zone~7 (partial load) also show consistently high performance. Errors that do occur are
almost entirely confined to adjacent zones. There is no evidence of widespread
misclassification where, for example, a full-load observation is labelled as standby.

\subsection{External Validation (2017)}
\label{subsec:ch3:results:ext}

On the 2017 external test set, RIPPER achieves 91.2\% accuracy and a macro-F1 of 0.914.
This is a small drop from the 92.6\%/0.916 on the 2016 validation set, indicating good
generalisation. The pattern of errors is similar to the internal validation: Zone~6
remains the most challenging zone, and misclassifications continue to be concentrated
near zone boundaries.

FOIL achieves 91.4\% accuracy on the 2017 external test, a drop of 0.9 percentage
points from 92.3\% on the internal validation. The macro-F1 decreases from 0.897 to
0.896. Both declines are smaller than those for RIPPER, which drops by 1.4 percentage
points in accuracy. The smaller decline for FOIL may be explained by the 2017 data
having a more balanced distribution of zone frequencies than the last 20\% of the 2016
dataset, which gives FOIL's more specific rules more opportunity to match correctly.

The external validation results confirm that both classifiers generalise well across
years. The learned rules capture stable physical relationships between the SCADA variables
and the operational zones, relationships that remain valid twelve months after the
training data was collected, without any retraining.

\subsection{Comparison of RIPPER and FOIL Rule Structures}
\label{subsec:ch3:results:rules}

Beyond accuracy, the structure of the learned rule sets shows important differences
between the two algorithms. For Zone~7 (partial load), RIPPER produces six rules. The
first four all combine torque-controlled rotor speed with medium wind speed and one of
several power output levels. The remaining two extend coverage to high-wind conditions
and high-power output at full-efficiency pitch angle, capturing observations near the
Zone~7/8 boundary. Together, the six rules cover Zone~7 using conditions that match the
known physics of partial load operation closely.

FOIL produces eleven separate rules for the same zone, each combining four conditions,
covering every combination of power level and pitch angle individually. While every
individual FOIL rule is physically interpretable, the number of rules makes the Zone~7
classification harder to summarise at a glance. RIPPER's more compact representation is
preferable in practice when a domain expert needs to read and check the classification
logic.

The compact structure of RIPPER rules is consistent across all eight zones: RIPPER
produces fewer rules per zone and each rule covers a larger part of the feature space.
This may explain why RIPPER maintains a small macro-F1 advantage over FOIL in the
external test (0.914 vs.\ 0.896): rules that cover more of the feature space are less
sensitive to small shifts in the data distribution between years.

\section{Discussion}
\label{sec:ch3:discussion}

The results show that physics-informed rule-based classifiers can reliably identify wind
turbine operational zones from SCADA data, reaching 91--93\% accuracy on both internal
validation and an independent external test set from a different year. The learned rules
are physically interpretable and can be checked against known turbine control logic.
This addresses Research Question~1 directly.

The main source of error is the transition zones, particularly Zone~6 (generator
engagement), which occupies a very narrow band of rotor speeds. The discretization of
continuous RPM values into bins means that observations near the Zone~5/6 or Zone~6/7
boundary can end up in the wrong bin due to the averaging effect of ten-minute data. This
is not a problem with the zone definition itself but a limitation of how the data is
represented for rule induction.

The comparison between RIPPER and FOIL shows that both are viable for this task. RIPPER
shows a higher macro-F1 on both evaluation sets (0.916 vs.\ 0.897 internally, 0.914
vs.\ 0.896 externally) and produces a more compact rule set. For a practical PHM
application where a domain expert needs to read and verify the classification rules,
RIPPER is the preferred choice. FOIL remains useful as an independent check: when RIPPER
and FOIL agree, there is high confidence in the prediction. When they disagree, the
disagreement itself signals that the observation is near a zone boundary and may need
closer inspection.

This study does not compare the rule-based methods against other interpretable
classifiers such as decision trees. That comparison is addressed in Chapter~4, where
supervised classifiers including Random Forests and Gradient Boosting are evaluated on
the same task using the zone labels produced by this study.

\section{Summary}
\label{sec:ch3:summary}

This chapter presented a physics-informed framework for defining and classifying nine
operational zones for wind turbine SCADA data, eight of which appear in the dataset. The
zone boundaries were derived from known turbine control parameters rather than from
statistical analysis. RIPPER achieved 92.6\% accuracy on internal validation and 91.2\%
on the 2017 external test. FOIL achieved 92.3\% and 91.4\% respectively. Both classifiers
generalise well across years and produce errors confined to adjacent zone boundaries,
reflecting the genuine difficulty of separating transition states rather than a failure
of the zone framework itself.

The zone labels produced by this framework and the RIPPER one-vs-rest probability outputs
form the foundation for Chapter~4, which compares DRL agents and supervised classifiers
on the same classification task.
