%!TEX root = thesis.tex
\chapter{Introduction}
\label{chap:introduction}

The global expansion of wind energy has been one of the defining energy trends
of the last two decades \citep{didier2022}. Offshore wind in particular has grown
rapidly, with turbines becoming larger and wind farms more complex. As turbines get larger and wind
farms become more complex, the challenge of keeping them running reliably and efficiently
has become more important and more difficult.

Every modern wind turbine is equipped with a Supervisory Control and Data Acquisition
(SCADA) system that records many measurements \citep{pandit2024}: wind speed, rotor speed, blade pitch angle,
generator power output, temperatures, humidity and much more. This data is the primary
window into the performance and health of the turbine. But the data is not always easy
to interpret. A reading of 500~kW on the power output sensor means something very
different if the turbine is ramping up from a standstill than if it has been running
steadily at partial load for hours. The same sensor value can indicate healthy operation,
a fault, or a planned control action, depending entirely on the broader context of what
the turbine is doing at that exact moment.

This thesis is about building that context. Specifically, it is about identifying which
of a set of well-defined operating states, referred to as operational zones, a wind
turbine is currently in. Getting this right is the first step for almost every downstream
monitoring and diagnostics task. This thesis develops and evaluates methods for this
classification problem using two different approaches: rule-based machine learning and
deep reinforcement learning.

\section{Wind Turbines and the Importance of Operational Context}
\label{sec:intro:context}

Wind turbines are complex pieces of machinery that operate across a wide range of
conditions. At lower wind speeds, the turbine sits in standby not producing anything.
As the wind picks up, the control system gradually spins up the rotor, the generator
engages and the turbine begins producing power. At higher wind speeds, the turbine
operates in a partial load state where the goal is to extract as much energy as possible
from the wind. Above a rated wind speed limit, the turbine has reached its maximum power
output and the control system is actively limiting power production by adjusting the blade
pitch angle. At extreme wind speeds, the turbine shuts down entirely to protect itself
from damage.

Each of these operating states, and the transitions between them, produces its own
pattern in the SCADA data. Wind speed, rotor RPM and blade pitch angle all behave
according to different physical laws in each state. In partial load, for example, the
power output follows a cubic relationship with wind speed and blade pitch is held at
zero degrees to maximise aerodynamic efficiency. In full load, the same power output
is held constant while blade pitch increases actively to limit rotor torque. A model
that does not know which of these two states the turbine is in will struggle to make
sense of either one.

This is the core problem that motivates this thesis. Before any other analysis of
SCADA data can be done reliably, before fault detection, before power curve modelling,
before degradation tracking, the operating state of the turbine must be identified.
This thesis calls this problem operational zone classification.

\section{SCADA Systems in Wind Turbine Monitoring}
\label{sec:intro:scada}

SCADA, as mentioned, stands for Supervisory Control and Data Acquisition. In a wind
turbine, the SCADA system serves two main roles: it controls the turbine (adjusting blade
pitch, managing the grid connection, controlling yaw to face the wind) and it records data
about how the turbine is performing. Most modern wind turbines store SCADA data as
ten-minute averages of each measured variable, alongside standard deviations and minimum
and maximum values over the same period. A typical wind turbine records a large number of variables \citep{pandit2024, pandit2019}.

The dataset used in this thesis consists of SCADA data provided by EDP (Energias de
Portugal) and covers two wind turbines at an offshore wind farm located off the west
coast of Africa. In total, 76 SCADA
variables are recorded. For the purposes of this thesis, four variables are central:
ambient wind speed (\texttt{Amb\_Wind\_Speed}), rotor angular speed
(\texttt{Rtr\_RPM\_Avg}), blade pitch angle (\texttt{Blades\_Pitch\_Angle}) and grid
power output (\texttt{Grid\_Prod\_Power}). These four variables describe the complete
aerodynamic and electrical state of the turbine and are sufficient to determine which
operating zone the turbine is in.

SCADA data presents a specific challenge: it is generated by a closed-loop control system.
This means the sensors do not record a passively observed physical process, they record
the result of the turbine actively managing itself. When wind speed increases, the control
system responds by adjusting blade pitch, which changes rotor speed, which changes the
power output. Interpreting this data correctly requires understanding the control logic,
not just the raw sensor values. Physics-informed approaches, like the zone definitions
used in this thesis, are important precisely because they encode this control logic
explicitly.

\section{Operational Zones: Definition and Importance}
\label{sec:intro:zones}

The concept of an operational zone, as used in this thesis, refers to a physically
well-defined state that a wind turbine can be in. Each zone corresponds to a specific
combination of wind conditions and control actions, and each has a clear physical
interpretation. The zone framework developed in the first study of this thesis \citep{indrekvam2026}
defines nine zones:

\begin{itemize}
  \item \textbf{Zone 1 - Safety Lockdown}
        (\texttt{Blades\_Pitch\_Angle} $\geq$ 86\textdegree):
        the turbine is intentionally prevented from producing power. Blade pitch is held
        at or above 86\textdegree, typically triggered by external constraints such as
        grid saturation or scheduled maintenance.

  \item \textbf{Zone 2 - Transition to or from Safety Lockdown}
        (\texttt{Blades\_Pitch\_Angle} $<$ 86\textdegree):
        blade pitch is actively adjusted as the turbine moves between operational zones
        and the lockdown state. Transitions span more than one SCADA sampling interval,
        producing intermediate readings in this zone.

  \item \textbf{Zone 3 - Standby (Low-Wind Shutdown)}
        (\texttt{Amb\_Wind\_Speed} $<$ 4\,m/s,
         \texttt{Blades\_Pitch\_Angle} $=$ 24\textdegree,
         \texttt{Rtr\_RPM} $=$ 0):
        wind speed is below the cut-in threshold, the turbine is fully braked, blade
        pitch is fixed at approximately 24\textdegree, and no power is produced.

  \item \textbf{Zone 4 - Transition to or from Standby}
        (\texttt{Amb\_Wind\_Speed} $<$ 4\,m/s,
         \texttt{Blades\_Pitch\_Angle} $\approx$ 24\textdegree,
         \texttt{Rtr\_RPM} $>$ 0):
        during startup or shutdown under low-wind conditions the brakes are released and
        the rotor begins to move while pitch remains near 24\textdegree. These are
        control-driven transient states rather than steady operation.

  \item \textbf{Zone 5 - Free Wheeling (Below Cut-In)}
        (\texttt{Amb\_Wind\_Speed} $<$ 4\,m/s,
         0\textdegree $<$ \texttt{Blades\_Pitch\_Angle} $<$ 24\textdegree,
         0 $<$ \texttt{Rtr\_RPM} $<$ 11):
        the rotor spins freely at low speed without generator engagement. Blade pitch
        gradually decreases toward 0\textdegree as wind speed increases.

  \item \textbf{Zone 6 - Generator Engagement (Below Cut-In)}
        (\texttt{Amb\_Wind\_Speed} $<$ 4\,m/s,
         \texttt{Blades\_Pitch\_Angle} $=$ 0\textdegree,
         11.0 $<$ \texttt{Rtr\_RPM} $\leq$ 11.2):
        the generator connects to the grid in a narrow rotor speed band of approximately
        11.0--11.2\,RPM. Rotor speed remains nearly constant while torque and power
        begin to increase.

  \item \textbf{Zone 7 - Partial Load Operation}
        (4 $\leq$ \texttt{Amb\_Wind\_Speed} $\leq$ 12\,m/s,
         \texttt{Blades\_Pitch\_Angle} $=$ 0\textdegree,
         11.2 $<$ \texttt{Rtr\_RPM} $<$ 14.8):
        between cut-in and rated wind speed, blade pitch is fixed at 0\textdegree and
        power output increases following the cubic power law. This is the most common
        operating state, accounting for approximately 51\% of observations in the dataset.

  \item \textbf{Zone 8 - Full Load Operation}
        (12 $<$ \texttt{Amb\_Wind\_Speed} $<$ 25\,m/s,
         \texttt{Blades\_Pitch\_Angle} $>$ 0\textdegree,
         11.8 $\leq$ \texttt{Rtr\_RPM} $<$ 15):
        above rated wind speed the turbine reduces aerodynamic efficiency to prevent
        overspeed and overloading. Blade pitch is actively increased while rotor speed
        and power output are regulated to approximately constant values
        (around 15\,RPM and 2000\,kW respectively).

  \item \textbf{Zone 9 - High-Wind Safety Shutdown}
        (\texttt{Amb\_Wind\_Speed} $>$ 25\,m/s):
        at extreme wind speeds the turbine automatically shuts down to prevent structural
        overload. Blades are pitched toward 90\textdegree and the rotor is braked to a
        stop. No observations corresponding to this zone were found in the dataset used
        in this thesis, so the classification problem covers Zones 1 to 8 in practice.
\end{itemize}

Getting zone classification right matters because the physical behaviour of the turbine
is fundamentally different in each zone. A condition monitoring system that does not
distinguish between zones could generate false alarms whenever the turbine moves between
them. Power curve models trained without zone separation could mix data from different
control regimes and might produce inaccurate results. Operational zone classification is
the foundation on which reliable wind turbine monitoring is built, and this thesis treats
it as its primary research problem.

\section{Prognostics, Health Management, and Digital Twins}
\label{sec:intro:phm}

Prognostics and Health Management (PHM) is the field concerned with monitoring the
health of industrial systems, predicting when they will fail, and planning maintenance
accordingly. For wind turbines, PHM is especially important because turbines are often
located in remote or offshore locations where unplanned maintenance is expensive and
operationally challenging \citep{tavner2012}. Being able to predict a fault days,
weeks or months in advance allows maintenance crews to plan visits efficiently and avoid
unexpected downtime.

A digital twin is a computational model of a physical asset that is continuously updated
with data from the real system. For a wind turbine, a digital twin might include models
of the power curve, the drivetrain dynamics, structural loads, and expected sensor
readings under normal operating conditions. When the real turbine deviates from what the
digital twin predicts, this signals a potential fault or degradation. The definition
adopted in this thesis, from the National Academies of Sciences, Engineering, and
Medicine \citeyearpar{nationalacademies2024}, describes a digital twin as \textit{``a
set of virtual information constructs that mimics the structure, context, and behaviour
of a natural, engineered, or social system; is dynamically updated with data from its
physical twin; has a predictive capability; and informs decisions to realize
value.''}

For a digital twin to work well, its models need to be context-aware. A power curve
model designed for Zone~7 (partial load) is not the right model for Zone~8 (full load),
even though both zones involve the turbine producing power. Operational zone
classification is therefore the necessary foundation for making every other model in the
digital twin context-aware. This thesis focuses specifically on this task.

\section{Approaches to Zone Classification}
\label{sec:intro:approaches}

Multiple approaches to classifying wind turbine operating states have been explored.
\citep{pandit2024} provides a comprehensive overview of SCADA-based methods for wind
turbine monitoring.
Physics-based threshold rules are the most direct and are already used in many SCADA
monitoring systems \citep{pandit2024}. However, hand-crafted rules require expert knowledge
and do not automatically handle sensor noise or transition states.

Rule-based machine learning algorithms, such as Repeated Incremental Pruning to Produce
Error Reduction (RIPPER) \citep{cohen1995} and First-Order Inductive Learner (FOIL)
\citep{quinlan1990}, can learn classification rules automatically from labelled data.
These methods produce explicit IF--THEN rules that can be read and verified by a domain
expert, which makes them attractive for safety-critical applications where interpretability
is important. The first study in this thesis uses these methods \citep{indrekvam2026}.

Supervised machine learning classifiers, such as Random Forests, Gradient Boosting and
neural networks, typically achieve higher accuracy than rule-based methods but produce
predictions without a human-readable explanation. Deep Reinforcement Learning (DRL) is a
further alternative that does not require labelled training data at all: a DRL agent
learns by receiving reward signals indicating whether its predictions are correct. This
makes DRL a candidate for settings where expert labelling is unavailable or where having a human expert manually label
thousands of readings is not practical.
The second study in this thesis evaluates all of these approaches side by side.

The closest existing work to this thesis focuses on SCADA-based analytics for wind turbine
performance monitoring \citep{pandit2024} and on DRL for wind turbine torque and pitch
control \citep{xie2023, didier2022}. However, detailed physics-consistent zone
classification covering transition states, and the systematic comparison of rule-based,
supervised, and DRL methods on the same classification problem, have not been addressed
in the literature. This gap is the starting point for the two studies in this thesis.

\section{Research Gap and Motivation}
\label{sec:intro:gap}

Despite a large body of work on wind turbine SCADA analysis \citep{pandit2024}, two
specific gaps motivate this thesis. First, most existing studies either use very simple
operating state definitions, separating only power production from standby \citep{manwell2009},
or use no state classification at all. A detailed, physics-consistent zone framework that
includes the critical transition states has not yet been addressed.

Second, no study has systematically compared rule-based methods, supervised classifiers,
and DRL for the same operational zone classification problem on the same dataset. Without
this comparison, practitioners have no clear guidance on which approach is most suitable
for their situation. This thesis addresses both gaps directly.

Taken together, these two gaps define the scope of this work. The overarching research problem is therefore: how can operational zones in a wind turbine
be identified reliably from SCADA data, and which machine learning approach is most suitable
depending on whether interpretability, accuracy, or the ability to adapt over time is the
priority?

\section{Research Questions}
\label{sec:intro:rq}

This thesis addresses the following research questions:

\begin{enumerate}
  \item How accurately and interpretably do physics-informed rule-based classifiers
        (RIPPER and FOIL) identify operational zones from wind turbine SCADA data,
        and how well do they generalise across different turbines and years?

  \item How much accuracy improvement do supervised machine learning classifiers
        (Random Forest, Gradient Boosting, MLP) achieve over rule-based methods for
        the same zone classification task, and what explains this improvement?

  \item How effectively do deep reinforcement learning agents (DQN, PPO, A2C) learn
        to classify operational zones, and under what deployment conditions are they
        preferable to supervised classifiers?

  \item What is the contribution of RIPPER-derived probability features to
        classification accuracy, compared to using only the four raw SCADA signals?

  \item To what extent do physics-based guardrails improve DRL agent accuracy and
        physical compliance, and how does this vary across different DRL algorithms?
\end{enumerate}

\section{Scope and Delimitations}
\label{sec:intro:scope}

This thesis uses SCADA data from two offshore wind turbines at a wind farm off the west
coast of Africa. Wind Turbine~7 provides the 2016 training and internal validation data.
Wind Turbine~11 from the same wind farm provides the 2017 external test data. No other
turbines, wind farms, or time periods are included. The findings should be interpreted as
results for this specific turbine type and data source and not generalised without further
validation.

The thesis does not evaluate true online learning. All DRL agents are trained in a batch
setting using historical data, not on a live turbine. The potential of DRL for continuous
adaptation during deployment is discussed but not tested. The thesis also does not
evaluate the downstream impact of zone classification on specific PHM tasks such as fault
detection or power curve modelling, and this is identified as an important direction for
future work.

\section{Research Method}
\label{sec:intro:method}

This thesis employs an empirical, experimental research approach structured as two
connected studies. No primary data was collected. All experiments use SCADA data provided
by EDP, described in Section~\ref{sec:intro:scada}. The dataset consists exclusively of
machine sensor readings with no personal or identifiable information, and no ethical
approval was required.

The literature review was conducted as a structured, targeted review across four thematic
areas: wind turbine aerodynamics and SCADA-based condition monitoring, prognostics and
health management for wind energy systems, rule-based machine learning, and deep
reinforcement learning for industrial applications. An initial structured search was
carried out using Google Scholar, IEEE Xplore, and Semantic Scholar with search terms
including ``wind turbine SCADA classification'', ``operational zone detection'',
``rule-based machine learning PHM'', ``RIPPER rule induction'', and ``deep reinforcement
learning industrial control''. This was complemented by supervisor-recommended
foundational works and forward and backward citation tracking from key papers.

Both empirical studies follow the same experimental protocol. Training and internal
validation use data from Wind Turbine~7 collected in 2016, split chronologically: the
first 80\% of the time series serves as the training set and the remaining 20\% as the
internal validation set. This chronological split preserves the temporal structure of the
data and evaluates generalisation in the direction of actual operational deployment.
External generalisation is evaluated on the full 2017 dataset from Wind Turbine~11, a
different turbine at the same wind farm, without any retraining or parameter adjustment.
This cross-turbine, cross-year evaluation tests how well the models work on data they
have never seen, collected from a different turbine in a different year.

All models are implemented in Python. Supervised classifiers and meta-models use
scikit-learn \citep{pedregosa2011}. DRL agents are implemented with Stable-Baselines3
\citep{raffin2021} and trained in a custom environment built with Gymnasium
\citep{towers2023}. All experiments use fixed random seeds and fully documented
hyperparameters detailed in Appendix~B, making results reproducible. The code is stored
in a private GitHub repository and will be made available upon reasonable request.

Performance is evaluated using overall accuracy and macro-averaged F1 score. Macro-F1
gives equal weight to all eight operational zones present in the dataset, which is
important given the strong class imbalance (Zone~7 accounts for approximately 51\% of
all readings). Per-zone precision, recall, F1-score, and support for the two
best-performing models are reported in Appendix~C.

\section{Thesis Structure}
\label{sec:intro:structure}

The remainder of this thesis is structured as follows.

Chapter~2 provides the technical background for both studies, covering wind turbine
aerodynamics and control, SCADA-based condition monitoring, digital twins, rule-based
machine learning, supervised classifiers, and deep reinforcement learning.

Chapter~3 presents the first study: physics-informed operational zone definition and
rule-based classification \citep{indrekvam2026}. It describes how the nine operational
zones were defined from turbine physics, how SCADA variables were discretized, and how
RIPPER and FOIL were trained and evaluated on both internal and external validation sets.

Chapter~4 presents the second study: a comparison of DRL agents and supervised
classifiers for the same zone classification task. It covers the feature representation,
DRL environment design, physics guardrails, full results across all models, and a
comparison of two deployment pipelines showing how DQN can reduce operator review
workload by approximately 70\%. This chapter is based on a companion manuscript prepared
for journal submission.

Chapter~5 brings the two studies together, interprets the key findings jointly, and
discusses what the results mean for the design of a PHM digital twin for wind turbines.

Chapter~6 states the conclusions, provides direct answers to the five research questions,
describes limitations and threats to validity, and proposes directions for future
research.

The thesis ends with a reference list and eight appendices: RIPPER rule sets (Appendix~A), model hyperparameters (Appendix~B),
per-zone classification reports (Appendix~C), a glossary of key terms (Appendix~D),
source code availability (Appendix~E), research data information (Appendix~F), and the
two published manuscripts on which Chapters~3 and~4 are based (Appendices~G and~H).
